---
title: "Reproducible Research Project using UCI Activity Recognition Data (Johns Hopkins Data Science Certification)"
author: "MD Alamgir Ph.D."
date: "August 29, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Project Description

Impute missing values and update the dataset

* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  

* Devise a strategy for filling in all of the missing values in the dataset. 
        - The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  

* Create a new dataset that is equal to the original dataset but with the missing data filled in.  

* Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

        - Do these values differ from the estimates from the first part of the assignment?
        
        - What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Load the data

```{r Read Data}
# Set working directory
setwd("C:/Users/MD/Documents")
# Data Source: UCI Activity Recognition database
# https://archive.ics.uci.edu/ml/datasets/Heterogeneity+Activity+Recognition
# Download the zipfile and extract it to the working directory.
# Load data to a data frame dfrr
dfrr <- read.csv("activity.csv")
```
##Total Number of Daily Steps

* Calculate the total number of steps taken per day  
* Make a histogram of total number of steps taken each day 
* Calculate and report the mean and the median of the total number of steps taken per day 

```{r totaldailysteps, histogram, mean, median}
# tds is total daily steps

tds <- tapply (dfrr$steps, dfrr$date,sum)

# create a histogram from tds slicing thedata into 5 bins, so breaks=5+1=6.
hist(tds, breaks=6, col="blue", xlab="Number of Daily Steps", ylab = "Frequency", main ="Frequency of Daily Steps")

# tdsmean = mean of total daily steps
# tdsmed  = median of total daily steps

tdsmean <- mean(tds, na.rm=TRUE)
tdsmean
tdsmed <- median(tds, na.rm=TRUE)
tdsmed

```
## Average Daily Activity Pattern

* Create a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
* Determine the 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps.

```{r time series plot and the 5-minute interval}
# mspi = mean steps perinterval
 mspi <- tapply(dfrr$steps, dfrr$interval, mean, na.rm=TRUE)

# plot mspi as time-series (type="l")


plot(mspi, type = "l", main= "Daily Average of # of Steps Taken per 5-minute Interval", xlab = "Interval Index #", ylab ="Average # of Steps", col="red")

```

```{r interval index having the maximum number of steps}
# interval index having the maximum number of steps
which(mspi %in% max(mspi))
